---
title: "Introduction to Data Visualization with ggplot2"
format:
  html:
    toc: true
author: "The Graph Courses Team"
---

# Introduction

Welcome to the world of data visualization with ggplot2! In this workshop, you'll learn how to create beautiful, informative visualizations using R's most popular plotting package. ggplot2 is built on the "Grammar of Graphics" - a systematic approach to creating plots by combining different components.

By the end of this workshop, you'll be able to:

1. Understand the basic structure of ggplot2 plots
2. Create scatter plots, bar charts, histograms, and box plots
3. Customize colors, themes, and labels
4. Use faceting to create multiple subplots
5. Apply best practices for effective data visualization

::: {.callout-note}
**Prerequisites:** This workshop assumes basic familiarity with R. If you're new to R, consider reviewing basic R syntax and data manipulation before proceeding.
:::

------------------------------------------------------------------------

# Part 1: Understanding the Grammar of Graphics

## The ggplot2 Philosophy

ggplot2 follows a layered approach to building plots. Every ggplot2 visualization consists of:

1. **Data**: The dataset you want to visualize
2. **Aesthetics (aes)**: How variables map to visual properties (x, y, color, size, etc.)
3. **Geometries (geom)**: The type of plot (points, bars, lines, etc.)
4. **Additional layers**: Statistics, scales, coordinate systems, themes, and facets

## Setting Up Your Environment

First, let's install and load the necessary packages:

```{r setup, eval=FALSE}
# Install packages if you haven't already
install.packages(c("ggplot2", "dplyr", "readr"))

# Load the packages
library(ggplot2)
library(dplyr)
library(readr)
```

## Your First ggplot2 Plot

Let's start with the built-in `mpg` dataset, which contains fuel economy data for different car models:

```{r first-plot, eval=FALSE}
# Load the data
data(mpg)

# Examine the structure
head(mpg)
str(mpg)

# Create a basic scatter plot
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point()
```

**What's happening here?**
- `ggplot()` initializes the plot with data and aesthetic mappings
- `aes(x = displ, y = hwy)` maps engine displacement to x-axis and highway mpg to y-axis
- `geom_point()` adds points to create a scatter plot
- The `+` operator adds layers to the plot

------------------------------------------------------------------------

# Part 2: Essential Plot Types

## Scatter Plots

Scatter plots are perfect for exploring relationships between two continuous variables:

```{r scatter-plots, eval=FALSE}
# Basic scatter plot
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point()

# Add color by categorical variable
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point()

# Add size aesthetic
ggplot(mpg, aes(x = displ, y = hwy, color = class, size = cty)) +
  geom_point()

# Add transparency to handle overplotting
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.6)
```

## Bar Charts

Bar charts are ideal for categorical data:

```{r bar-charts, eval=FALSE}
# Simple bar chart (count)
ggplot(mpg, aes(x = class)) +
  geom_bar()

# Horizontal bar chart
ggplot(mpg, aes(x = class)) +
  geom_bar() +
  coord_flip()

# Stacked bar chart
ggplot(mpg, aes(x = class, fill = drv)) +
  geom_bar()

# Side-by-side bars
ggplot(mpg, aes(x = class, fill = drv)) +
  geom_bar(position = "dodge")

# Bar chart with calculated values
mpg_summary <- mpg %>%
  group_by(class) %>%
  summarise(avg_hwy = mean(hwy))

ggplot(mpg_summary, aes(x = class, y = avg_hwy)) +
  geom_col()
```

## Histograms and Density Plots

For exploring distributions of continuous variables:

```{r histograms, eval=FALSE}
# Basic histogram
ggplot(mpg, aes(x = hwy)) +
  geom_histogram()

# Adjust bin width
ggplot(mpg, aes(x = hwy)) +
  geom_histogram(binwidth = 2)

# Density plot
ggplot(mpg, aes(x = hwy)) +
  geom_density()

# Multiple densities by group
ggplot(mpg, aes(x = hwy, fill = drv)) +
  geom_density(alpha = 0.7)
```

## Box Plots

Box plots show distributions and identify outliers:

```{r box-plots, eval=FALSE}
# Basic box plot
ggplot(mpg, aes(x = class, y = hwy)) +
  geom_boxplot()

# Add individual points
ggplot(mpg, aes(x = class, y = hwy)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.6)

# Violin plots (combination of box plot and density)
ggplot(mpg, aes(x = class, y = hwy)) +
  geom_violin()
```

------------------------------------------------------------------------

# Part 3: Customizing Your Plots

## Adding Trend Lines

```{r trend-lines, eval=FALSE}
# Linear trend line
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth(method = "lm")

# Smooth trend line (loess)
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth()

# Remove confidence interval
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

## Labels and Titles

```{r labels, eval=FALSE}
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point() +
  labs(
    title = "Fuel Efficiency vs Engine Size",
    subtitle = "Highway MPG decreases with larger engines",
    x = "Engine Displacement (liters)",
    y = "Highway Miles per Gallon",
    color = "Vehicle Class",
    caption = "Data source: EPA fuel economy data"
  )
```

## Color and Fill Scales

```{r colors, eval=FALSE}
# Manual colors
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point() +
  scale_color_manual(values = c("red", "blue", "green", "purple", 
                               "orange", "brown", "pink"))

# ColorBrewer palettes
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point() +
  scale_color_brewer(type = "qual", palette = "Set2")

# Viridis color scale (colorblind-friendly)
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point() +
  scale_color_viridis_d()
```

## Themes

```{r themes, eval=FALSE}
# Built-in themes
p <- ggplot(mpg, aes(x = displ, y = hwy)) + geom_point()

p + theme_minimal()
p + theme_classic()
p + theme_dark()
p + theme_void()

# Custom theme modifications
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point() +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

------------------------------------------------------------------------

# Part 4: Faceting - Multiple Plots

Faceting allows you to create multiple subplots based on categorical variables:

```{r faceting, eval=FALSE}
# Facet wrap - creates a grid of plots
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_wrap(~ class)

# Facet wrap with custom number of columns
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_wrap(~ class, ncol = 3)

# Facet grid - creates a matrix of plots
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_grid(drv ~ cyl)

# Free scales for better visibility
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_wrap(~ class, scales = "free")
```

------------------------------------------------------------------------

# Part 5: Working with Real Data

Let's work with a more complex dataset to practice our skills. We'll create a sample dataset about student performance:

```{r sample-data, eval=FALSE}
# Create sample student performance data
set.seed(123)
students <- data.frame(
  student_id = 1:200,
  study_hours = round(rnorm(200, mean = 15, sd = 5), 1),
  previous_grade = sample(c("A", "B", "C", "D"), 200, 
                         replace = TRUE, prob = c(0.2, 0.3, 0.3, 0.2)),
  subject = sample(c("Math", "Science", "English", "History"), 200, replace = TRUE),
  final_score = round(pmax(0, pmin(100, 
    50 + 2 * study_hours + 
    ifelse(previous_grade == "A", 15, 
           ifelse(previous_grade == "B", 10, 
                  ifelse(previous_grade == "C", 5, 0))) + 
    rnorm(200, 0, 8))), 1)
)

# Save the data
write_csv(students, "data/student_performance.csv")

# Examine the data
head(students)
summary(students)
```

## Comprehensive Analysis

```{r analysis, eval=FALSE}
# 1. Distribution of study hours
ggplot(students, aes(x = study_hours)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Study Hours",
       x = "Study Hours per Week",
       y = "Number of Students") +
  theme_minimal()

# 2. Relationship between study hours and final score
ggplot(students, aes(x = study_hours, y = final_score)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Study Hours vs Final Score",
       x = "Study Hours per Week",
       y = "Final Score") +
  theme_minimal()

# 3. Performance by previous grade
ggplot(students, aes(x = previous_grade, y = final_score, fill = previous_grade)) +
  geom_boxplot() +
  scale_fill_brewer(type = "qual", palette = "Set3") +
  labs(title = "Final Score by Previous Grade",
       x = "Previous Grade",
       y = "Final Score",
       fill = "Previous Grade") +
  theme_minimal()

# 4. Subject performance comparison
ggplot(students, aes(x = subject, y = final_score, color = subject)) +
  geom_jitter(width = 0.2, alpha = 0.6) +
  geom_boxplot(alpha = 0.3) +
  labs(title = "Performance by Subject",
       x = "Subject",
       y = "Final Score") +
  theme_minimal() +
  theme(legend.position = "none")

# 5. Complex faceted plot
ggplot(students, aes(x = study_hours, y = final_score, color = previous_grade)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ subject) +
  scale_color_brewer(type = "qual", palette = "Set1") +
  labs(title = "Study Hours vs Final Score by Subject and Previous Grade",
       x = "Study Hours per Week",
       y = "Final Score",
       color = "Previous Grade") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))
```

------------------------------------------------------------------------

# Part 6: Best Practices and Tips

## Do's and Don'ts

### ✅ Do:
- Choose appropriate plot types for your data
- Use clear, descriptive labels and titles
- Consider colorblind-friendly palettes
- Keep plots simple and focused
- Use consistent styling across related plots

### ❌ Don't:
- Overload plots with too much information
- Use 3D effects or unnecessary decorations
- Choose misleading scales or axes
- Use too many colors or patterns
- Forget to label axes and legends

## Common ggplot2 Patterns

```{r patterns, eval=FALSE}
# Save a plot object for reuse
base_plot <- ggplot(mpg, aes(x = displ, y = hwy))

# Add different geoms to the same base
base_plot + geom_point()
base_plot + geom_smooth()
base_plot + geom_point() + geom_smooth()

# Reorder factors for better visualization
mpg_reordered <- mpg %>%
  mutate(class = fct_reorder(class, hwy, median))

ggplot(mpg_reordered, aes(x = class, y = hwy)) +
  geom_boxplot()

# Create publication-ready plots
final_plot <- ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point(size = 2, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  scale_color_brewer(type = "qual", palette = "Set2") +
  labs(
    title = "Fuel Efficiency Decreases with Engine Size",
    subtitle = "Highway MPG vs Engine Displacement for Different Vehicle Classes",
    x = "Engine Displacement (liters)",
    y = "Highway Miles per Gallon",
    color = "Vehicle Class",
    caption = "Data: EPA fuel economy dataset"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12, color = "gray60"),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

# Save the plot
ggsave("fuel_efficiency_plot.png", final_plot, width = 10, height = 6, dpi = 300)
```

------------------------------------------------------------------------

# Part 7: Hands-On Exercises

## Exercise 1: Basic Plotting
Using the `mpg` dataset:
1. Create a scatter plot of city mpg (`cty`) vs highway mpg (`hwy`)
2. Color the points by fuel type (`fl`)
3. Add appropriate labels and a title

## Exercise 2: Categorical Data
Using the `mpg` dataset:
1. Create a bar chart showing the count of vehicles by manufacturer
2. Rotate the x-axis labels for better readability
3. Color the bars by drive type (`drv`)

## Exercise 3: Distributions
Using the student performance data:
1. Create overlapping density plots of final scores by subject
2. Use different colors for each subject
3. Add a vertical line showing the overall mean score

## Exercise 4: Advanced Visualization
Combine multiple techniques:
1. Create a faceted plot showing the relationship between study hours and final score
2. Facet by subject
3. Color points by previous grade
4. Add trend lines for each previous grade group
5. Apply a clean theme and proper labels

------------------------------------------------------------------------

# Part 8: Going Further

## Additional Resources

- **Official ggplot2 documentation**: [ggplot2.tidyverse.org](https://ggplot2.tidyverse.org/){target="_blank"}
- **R Graphics Cookbook**: Comprehensive guide to ggplot2
- **Data Visualization: A Practical Introduction**: By Kieran Healy
- **ggplot2 extensions**: [exts.ggplot2.tidyverse.org](https://exts.ggplot2.tidyverse.org/){target="_blank"}

## Advanced Topics to Explore

1. **Interactive plots** with `plotly`
2. **Animation** with `gganimate`
3. **Maps** with `ggmap` and `sf`
4. **Custom themes** and branding
5. **Statistical transformations** and custom geoms

## Next Steps

1. Practice with your own datasets
2. Explore different geom types and statistical transformations
3. Learn about the underlying grid graphics system
4. Contribute to the ggplot2 community

------------------------------------------------------------------------

# Submission

For this workshop, create a portfolio of visualizations that demonstrates your understanding of ggplot2:

1. **Create 5 different plots** using the techniques learned in this workshop
2. **Include variety**: Use different geom types, color schemes, and themes
3. **Document your code**: Add comments explaining your choices
4. **Write brief interpretations**: What story does each plot tell?

Submit your R script or R Markdown file containing:
- Your code for all 5 plots
- Brief explanations of what each plot shows
- Any insights or patterns you discovered in the data

**Bonus**: Create one "publication-ready" plot with custom styling, proper labels, and professional appearance.

------------------------------------------------------------------------

*Happy plotting! Remember, the best way to learn ggplot2 is through practice. Don't be afraid to experiment with different combinations of aesthetics, geoms, and themes.*


